\textit{MSF: Modulated Sub-graph Finder}

### 2018-02-20 MSF_Paper_FA.tex
nothing
Added 6hpi network

\title{\textit{MSF: Modulated Sub-path Finder} }

@@

\begin{abstract}
Differentially expressed genes play an important role in giving
insights to a phenotype. These DEGs are then used to identify the
pathways altered during changing biological conditions. We developed a
novel approach to find the maximal significantly dis-regulated
sub-paths or cluster of genes from the cell signaling network, giving
these sub-paths an overall significance of modulation by combining the
individual p-values of the genes.
\end{abstract}

@@

Differentially expressed genes, Pathway analysis, Combining \textit{p}-value, Cell signaling network, Modulated Sub-paths

@@

The existing pathway-based analysis approaches use different research
designs, which can be categorized into ORA (Over-representation
analysis), FCS (Functional class scoring) and pathway topology based
methods. ORA is the first and the most basic method of pathway
analysis. It uses a user defined cut-off for the log-fold change and
p-value from the DGEA (most commonly using absolute log-fold change
$\geq$ 2 and \textit{p}-value $\leq$ 0.05) to define a list of
differentially expressed genes. Subsequently, sets of genes associated
with annotated pathways are tested for being overrepresented in the
set of differentially expressed genes. To this end, hypergeometric
distribution, chi-square tests, binomial probability or the Fisher’s
exact test are used. Thereby the information of the topology of the
pathways are ignored~\cite{Bayer}. Futhermore, ORA assumes the
pathways are independent of each other and ignores the fact that
biological pathways cross-talk and overlap~\cite{Khatri2012,Campos}.

@@

On these grounds we propose a novel approach to make use of the rich
pathway annotation resources available to gain additional functional
insights from basic DGEA. To this, we
start with the presupposition that expression of neighboring genes
within a functional pathway are not independent from each
other. Rather, they are often regulating each others expression or are
part of the same regulon~\cite{Michalak}. Second, we
understand that the categorization of links between genes into labeled
pathways is often an arbitrary one, given the extensive cross talk
between different pathways. Although this categories have proven to be
useful in many situations, they force a certain perspective onto the
interpretation of novel data. Based on this two principles, we aim to
find sub-modules within predefined networks which exhibits as a whole
significant differential expression changes. As input information on
functional links between genes, provided by e.g. KEGG or Reactome, and
information on the differential expression status of single genes,
resulting from a DGEA, are required. As a result the analysis returns
sub-modules and their joint confidence scores, reflecting how the
perturbation is migrated through the network. Furthermore, the entry
points of perturbation in the networks and overlap with conventional
pathway categories are returned. All of this can be helpful to
understand cause end effect of a stimulus and might inform about
potential points of intervention. The proposed algorithm was
implemented as a java program, which was named Modulated SubPath
Finder (\texttt{MSF}).

\section*{Methods}
\texttt{MSF} is developed as a novel heuristic approach in Java to
find the modulated sub-paths of gene interactions from the cell
signaling network. The network has nodes corresponding to genes and
edges representing their relationship. The pipeline of identification
of modulated sub-paths from a network by \texttt{MSF} is depicted in
Fig (1).

\texttt{MSF} uses the individual gene's \textit{p}-values generated
from the DGEA. The \textit{p}-value
expresses the probability that the hypothesis of unmodified gene
expression can be rejected for a given statistical model. To find
significantly modulated sub-paths individual \textit{p}-values of the
vicinal genes in the global network are combined into a single
combined \textit{p}-value, using a statistical method for combining
dependent \textit{p}-values described by~\cite{Hartung}. Using the
inverse normal method, individual gene \textit{p}-values $p_i$ are
first transformed to its corresponding normal score $t_i$,

@@

followed by correction of the correlation \textbf{k}. The normal
scores, correlations and the correction of correlation are applied to
the inverse normal function to calculate the individual
\textit{p}-value for a sub-path \textit{Cp}. \newline

@@

Lambda $\lambda$ are the weights for each gene, For the moment equal
weights are given to all genes and \textbf{k} used is 0.2 as used by authors. An
overall \textit{p}-value of a sub-path will express the significance
of all genes in the sub-path being modulated together.

To reduce the complexity to score all possible connected sub-modules
\texttt{MSF} applies a three steps heuristic as described in the
following.\newline

@@

\textbf{Initial Modulated Sub-paths}

\texttt{MSF} constructs the first sub-path starting with the genes
associated with the lowest (most significant) \textit{p}-value from
the network. It tries to extend the
sub-path to the most significant neighboring gene. A single combined
\textit{p}-value is calculated for the two genes. If the combined
\textit{p}-value is smaller than the minimal individual gene's
\textit{p}-values, the extended sub-path is accepted. Then the next most significant neighboring gene is added to the sub-path and the combined \textit{p}-value is calculated, if the new combined p-value of three genes is smaller than the combined p-value of the first two genes,the extended sub-path is accepted. This step is iteratively repeated until no further extension is accepted. In this
case the process starts over with all remaining genes not yet in a
significant modulated sub-path. This step identifies all the trivial
sub-paths that are modulated in the whole network~(Fig.~1b).\newline

\textbf{Extending Modulated Sub-paths}

In the next step the initial modulated sub-paths are used to check if
they could further be extended beyond the immediate neighborhood.
This is done by testing all possible extension up to N genes for
all genes in the sub-path. Again, this step is iteratively repeated until no
further genes are added to the significant differentially expressed
sub-paths. This steps bridges small gaps of genes without a clear
differential signal in the DGEA~(Fig.~1c).\newline

\textbf{Merging Modulated Sub-paths}

After detection and extension of the modulated sub-paths, they are
tested if combined sub-path score is better than on their own. The most significant gene interaction of the first gene of a sub-path is taken and checked if it interacts with a gene in the second sub-path. If the two
paths merge with the connector of at most N genes and the
combined \textit{p}-value of the merged sub-path including the
bridging genes in between is less than the individual
\textit{p}-values of the two sub-paths, the two sub-paths are merged
together to one big modulated sub-module~(Fig.~1d). This step is repeated iteratively until no sub-paths could be merged to the sub-module.\newline

@@

In a last post processing step \texttt{MSF} identifies the trigger
points of the modulated sub-path. These trigger genes are the sources
of the sub-modules with only outgoing edges. These genes can be
interpreted as the entry points of perturbation from where the
stimulus causes downstream effects. In the same spirit the most
downstream genes of the modulated sub-path are identified and defined
as sinks. Sinks can be interpreted as the effectors where the
integrated information within the signal transduction network is set
to action.

@@

To demonstrate the usefulness \texttt{MSF} is applied to a RNAseq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus~\cite{Olejnik} (GSE84188). Ebola Virus (EBOV)
belongs to the Filoviridea family; filamentous, enveloped and single
stranded RNA viruses. EBOV causes hemorrhagic fever in humans,
inducing the host innate and adaptive immune response to be unable to
control virus infection~\cite{Prins}. Until now there are no approved
antiviral drugs for the treatment of Ebola virus infection~\cite{Konde,Rhein}. 
The initial targets of EBOV are the macrophages and
dendritic immune cells~\cite{Falasca,Rhein}. Ebola Virus inhibits the critical
innate immune response of the host, which includes the activation of
alpha/beta interferon (IFN-$\alpha /
\beta$)~\cite{Prins,Konde,Cardenas}. The EBOV viral protein VP35
targets the host type~I interferon IFN-$\alpha / \beta$ to block the
early antiviral immune response
~\cite{Prins,Konde,Falasca,Cardenas,Olejnik}. It has been proposed
that IFN-$\alpha / \beta$ could be used through clinical trials to
design antiviral drug against Ebola. The aim of testing the EBOV infection data with \texttt{MSF} was to  identify the modulated sub-paths and check if \texttt{MSF} is able to identify the IFN-$\alpha / \beta$ gene as one of the sources.

The EBOV infection data has three time-points six hour (6hpi), one day (1dpi) and two day (2dpi) post infection. At 6hpi, five modulated sub-modules were identified. Most of the genes part of the sub-moldules were chemokines (CXCL10, CCL8) and Interleukin genes (IL6, IL27, IL23) which serve as a subset of CD4+ T
cells. IFNB1  and IFNA1 were both identified as two of the sources. Most of the sources identified by \texttt{MSF} were type~I interferon induced genes. At 1dpi nine modulated sub-modules were identified, IFNA1 was identified as one of the sources. For the last time-point again IFNB1 and IFNA1 were identified as the two sources out of the possible sources from seven sub-modules identified.

As stated earlier IFN-$\alpha / \beta$ could be one of the target
genes of Ebola infection. We were able to successfully identify IFNA1
as a source in all three time-points and INFB1 in two of the time-points. Although IFNA1 and IFNB1 were two of the most
significant gene in the later time points, \texttt{MSF} was able to detect them as
a source in the early time-point when the genes were not significant
based on the individual DGEA alone. These sources can help the biologist as the starting point of clinical testing for drugs and vaccines against an infection.

\subsection*{Benchmark}

To better understand the functional role of the sub-modules identified by  \texttt{MSF}, we linked the  sub-modules to the predefined KEGG pathways and Reactome
pathways through gene enrichment analysis. We used Cytoscape~\cite{Cyto} plugins to identify the enriched patwhays, CytoKegg~\cite{Cytokegg} for KEGG and Reactome FI~\cite{Reactome} for Reactome. We compared the pathways enriched in \texttt{MSF} identified sub-modules to SPIA~\cite{Tarca} for KEGG and Reactome pathway enrichment analysis.

At 6hpi Cytokegg identified the sub-module genes to be enriched in toll-like receptor signaling, TNF signaling, IL-17 signaling and NF-kappa signaling pathways, as also stated by~\cite{Olejnik}. The earliest response EBOV
induces on  cytokine is via TLR4-mediated signaling~\cite{Olejnik}, Gene enrichment analysis of sub-module genes at 6hpi showed toll-like receptor signaling being most significantly dis-regulated pathway, this important signaling pathway was not identified by SPIA along with TNF signaling and IL-17 signaling (Table 1). On the later time-points \texttt{MSF} and SPIA showed consensus on the important pathways e.g Chemokine signaling, Cytokine-Cytokine receptor interaction, NF-kappa B signaling ,Cytosolic DNA-sensing  and RIG-I-like receptor signaling as being dis-regulated supplement (Table 3,4). SPIA identifies Cytokine-Cytokine receptor interaction and Chemokine signaling on the top for all the time
points and also shows apoptosis at the earliest time-point, where infection
just started.

Likewise we compared gene enrichment results of sub-module genes identified by \texttt{MSF} using Reactome FI to the Reactome pathway enrichment analysis results. Although MSF and Reactome show good agreement on the dis-regulated pathways e.g Signaling by Interleukins, Interleukin-10 signaling, Cytokine Signaling in Immune system, Chemokine receptors bind chemokines, RIG-I/MDA5 mediated induction of IFN-alpha/beta pathways, NF-kB activation, Reactome as well fails to identify the Toll-like recpetor signaling. On the other hand, \texttt{MSF} could not identify the dis-regulation of phases of cell cycle. This shows that \texttt{MSF} is able to identify the modulated genes in the sub-modules at the earliest time-point of infection even though when the signal transduction was weak.

@@

To show the robustness and stability of our method, Poisson
distributed noise was added to the sample read counts. 
Then DGEA was carried out on the disturbed data with the
same parameters as for the native data, followed by analysis with
\texttt{MSF}. Using this criteria, the goal was to spot almost
identical modulated sub-paths in the real and noisy data.

@@

	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=10.0cm]{Documents/PhDThesis/MSFPaper/Tgfb1} 
		\caption{}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=10.0cm]{Documents/PhDThesis/MSFPaper/TGFBstep1}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=10.0cm]{Documents/PhDThesis/MSFPaper/TGFBstep2} 
		\caption{}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=10.0cm]{Documents/PhDThesis/MSFPaper/TGFBstep3}
		\caption{}
	\end{subfigure}
	\caption{Overview of \texttt{MSF} process showing the steps to identify the modulated sub-paths. (a) showing a network of genes with green nodes as genes with \textit{p}-values $< $ 0.05, and blue nodes are genes with \textit{p}-values $>$ 0.05, (b) the genes circled red show the two initial modulated sub-paths, sub-path1 found one with M,K,C and the second modulated sub-path2 S,V,T, (c) shows the modulated sub-path1 being extended to M,K,C,A,D, (d) shows both the modulated sub=paths merged with the addition of 3 genes Z,W,P.}
\end{figure*}

\begin{table*}[]
	\centering
	\caption{Comparison of gene enrichment analysis of MSF 6hpi sub-modules with SPIA pathway analysis}
	\label{Comparison of gene enrichment analysis of MSF 6hpi sub-modules with SPIA pathway analysis}
	\begin{tabular}{|l|l|}
		\hline
		\textbf{MSF Sub-module Gene Enrichment}              & \textbf{SPIA Pathway Analysis}            \\ \hline
		Toll-like receptor signaling pathway                 & Cytokine-cytokine receptor interaction    \\ \hline
		TNF signaling pathway                                & Chemokine signaling pathway               \\ \hline
		Prion diseases                                       & Apoptosis                                 \\ \hline
		Pertussis                                            & Taste transduction                        \\ \hline
		Type II diabetes mellitus                            & Influenza A                               \\ \hline
		Thyroid cancer                                       & NF-kappa B signaling pathway              \\ \hline
		Toxoplasmosis                                        & Type II diabetes mellitus                 \\ \hline
		Prolactin signaling pathway                          & Osteoclast differentiation                \\ \hline
		Cytosolic DNA-sensing pathway                        & Cytosolic DNA-sensing pathway             \\ \hline
		Chagas disease (American trypanosomiasis)            & Natural killer cell mediated cytotoxicity \\ \hline
		RIG-I-like receptor signaling pathway                & Measles                                   \\ \hline
		Th1 and Th2 cell differentiation                     & Amyotrophic lateral sclerosis (ALS)       \\ \hline
		Chemokine signaling pathway                          & Salivary secretion                        \\ \hline
		IL-17 signaling pathway                              & HTLV-I infection                          \\ \hline
		AGE-RAGE signaling pathway in diabetic complications & MAPK signaling pathway                    \\ \hline
		Measles                                              & Calcium signaling pathway                 \\ \hline
		Jak-STAT signaling pathway                           & Notch signaling pathway                   \\ \hline
		Leishmaniasis                                        & Graft-versus-host disease                 \\ \hline
		Endometrial cancer                                   & Maturity onset diabetes of the young      \\ \hline
		ECM-receptor interaction                             & Herpes simplex infection                  \\ \hline

	\end{tabular}
\end{table*}

\begin{table*}[]
	\centering
	\caption{Comparison of gene enrichment analysis of MSF 6hpi sub-modules with Reactome pathway enrichment analysis}
	\label{my-label}
	\begin{tabular}{|l|l|}
		\hline
		\textbf{MSF Sub-module Gene Enrichment}          & \textbf{Reactome Pathway Enrichment Analysis}   \\ \hline
		Signaling by Interleukins                        & Interferon alpha/beta signaling                 \\ \hline
		Interleukin-10 signaling                         & Interferon Signaling                            \\ \hline
		Cytokine Signaling in Immune system              & Interleukin-10 signaling                        \\ \hline
		Interleukin-6 family signaling                   & Cytokine Signaling in Immune system             \\ \hline
		Growth hormone receptor signaling                & Signaling by Interleukins                       \\ \hline
		Chemokine receptors bind chemokines              & Interferon gamma signaling                      \\ \hline
		Interleukin-4 and 13 signaling                   & Interleukin-4 and 13 signaling                  \\ \hline
		Signalling by NGF                                & Chemokine receptors bind chemokines             \\ \hline
		IL-6-type cytokine receptor ligand interactions  & RIG-I/MDA5 mediated induction of IFN-alpha/beta \\ \hline
		GPVI-mediated activation cascade                 & Negative regulators of RIG-I/MDA5 signaling     \\ \hline
		MyD88 dependent cascade initiated on endosome    & Nucleotide-binding domain                       \\ \hline
		Toll Like Receptor 7/8 (TLR7/8) Cascade          & NF-kB activation through FADD/RIP-1 pathway     \\ \hline
		Activated TLR4 signalling                        & Ovarian tumor domain proteases                  \\ \hline
		CD28 dependent PI3K/Akt signaling                & GPCR ligand binding                             \\ \hline
		Toll Like Receptor 9 (TLR9) Cascade              & Interleukin-1 processing                        \\ \hline
		CD28 co-stimulation                              & TRAF3-dependent IRF activation pathway          \\ \hline
		NGF signalling via TRKA from the plasma membrane & Interleukin-6 family signaling                  \\ \hline
		Toll Like Receptor 3 (TLR3) Cascade              & Class A/1 (Rhodopsin-like receptors)            \\ \hline
		TRIF-mediated TLR3/TLR4 signaling                & Inflammasomes                                   \\ \hline
		MyD88-independent TLR3/TLR4 cascade              & Interleukin-1 signaling                         \\ \hline
	\end{tabular}
\end{table*}

\begin{figure*}[th]
	\includegraphics[width=16cm]{AllDEGSources}
	\caption{}
	\label{fig:alldegsources}
\end{figure*}

\begin{figure*}[p]
	\centering
	\includegraphics[width=14cm]{"SubPath Robustness 6H"}
\caption{}
\label{fig:subpath-robustness-6h}
\includegraphics[width=14cm]{Betweeness}
\caption{}
\label{fig:betweeness}
\end{figure*}

>>>>

\title{\textit{MSF: Modulated Sub-graph Finder} }

@@

\begin{abstract}
Differentially expressed genes play an important role in giving
insights to a phenotype. These DEGs are then used to identify the
pathways altered during changing biological conditions. We developed a
novel approach to find the maximal significantly dis-regulated
sub-graphs or cluster of genes from the cell signaling network, giving
these sub-graphs an overall significance of modulation by combining the
individual \textit{p}-values of the genes.
\end{abstract}

@@

Differentially expressed genes, Pathway analysis, Combining \textit{p}-value, Cell signaling network, Modulated Sub-graphs

@@

The existing pathway-based analysis approaches use different research
designs, which can be categorized into ORA (Over-representation
analysis), FCS (Functional class scoring) and pathway topology based
methods. ORA is the first and the most basic method of pathway
analysis. It uses a user defined cut-off for the log-fold change and
\textit{p}-value from the DGEA (most commonly using absolute log-fold change
$\geq$ 2 and \textit{p}-value $\leq$ 0.05) to define a list of
differentially expressed genes. Subsequently, sets of genes associated
with annotated pathways are tested for being overrepresented in the
set of differentially expressed genes. To this end, hypergeometric
distribution, chi-square tests, binomial probability or the Fisher’s
exact test are used. Thereby the information of the topology of the
pathways are ignored~\cite{Bayer}. Futhermore, ORA assumes the
pathways are independent of each other and ignores the fact that
biological pathways cross-talk and overlap~\cite{Khatri2012,Campos}.

@@

On these grounds we propose a novel approach to make use of the rich
pathway annotation resources available to gain additional functional
insights from basic DGEA. To this, we
start with the presupposition that expression of neighboring genes
within a functional pathway are not independent from each
other. Rather, they are often regulating each others expression or are
part of the same regulon~\cite{Michalak}. Second, we
understand that the categorization of links between genes into labeled
pathways is often an arbitrary one, given the extensive cross talk
between different pathways. Although this categories have proven to be
useful in many situations, they force a certain perspective onto the
interpretation of novel data. Based on this two principles, we aim to
find sub-graphs within predefined networks which exhibits as a whole
significant differential expression changes. As input information on
functional links between genes, provided by e.g. KEGG or Reactome, and
information on the differential expression status of single genes,
resulting from a DGEA, are required. As a result the analysis returns
sub-graphs and their joint confidence scores, reflecting how the
perturbation is migrated through the network. Furthermore, the entry
points of perturbation in the networks and overlap with conventional
pathway categories are returned. All of this can be helpful to
understand cause end effect of a stimulus and might inform about
potential points of intervention. The proposed algorithm was
implemented as a java program, which was named Modulated Sub-graph
Finder (\texttt{MSF}).

\section*{Methods}
\texttt{MSF} is developed as a novel heuristic approach in Java to
find the modulated sub-graphs of gene interactions from the cell
signaling network. \texttt{MSF} does not work on predefined pathways but builds up the sub-graph from scratch using the whole cell signaling network. The network has nodes corresponding to genes and
edges representing their relationship. The pipeline of identification
of modulated sub-graphs from a network by \texttt{MSF} is depicted in
Fig (1).

\texttt{MSF} uses the individual gene's \textit{p}-values generated
from the DGEA. The \textit{p}-value
expresses the probability that the hypothesis of unmodified gene
expression can be rejected for a given statistical model. To find
significantly modulated sub-graphs individual \textit{p}-values of the
vicinal genes in the global network are combined into a single
combined \textit{p}-value, using a statistical method for combining
dependent \textit{p}-values described by~\cite{Hartung}. Using the
inverse normal method, individual gene \textit{p}-values $p_i$ are
first transformed to its corresponding normal score $t_i$,

@@

followed by correction of the correlation \textbf{k}. The normal
scores, correlations and the correction of correlation are applied to
the inverse normal function to calculate the individual
\textit{p}-value for a sub-graph \textit{Cp}. \newline

@@

Lambda $\lambda$ are the weights for each gene, For the moment equal
weights are given to all genes and \textbf{k} used is 0.2 as used by authors. An
overall \textit{p}-value of a sub-graph will express the significance
of all genes in the sub-graph being modulated together.

To reduce the complexity to score all possible connected sub-graphs
\texttt{MSF} applies a three steps heuristic as described in the
following.\newline

@@

\textbf{Initial Modulated sub-graphs}

\texttt{MSF} constructs the first sub-graph starting with the genes
associated with the lowest (most significant) \textit{p}-value from
the network. It tries to extend the
sub-graph to the most significant neighboring gene. A single combined
\textit{p}-value is calculated for the two genes. If the combined
\textit{p}-value is smaller than the minimal individual gene's
\textit{p}-value, the extended sub-graph is accepted. Then the next most significant neighboring gene is added to the sub-graph and the combined \textit{p}-value is calculated, if the new combined \textit{p}-value of three genes is smaller than the combined \textit{p}-value of the first two genes,the extended sub-graph is accepted. This step is iteratively repeated until no further extension is accepted. In this case the process starts over with all remaining genes not yet in a
significant modulated sub-graph. This step identifies all the trivial
sub-graphs that are modulated in the whole network~(Fig.~1b).\newline

\textbf{Extending Modulated sub-graphs}

In the next step initial modulated sub-graphs are used to check if
they could further be extended beyond the immediate neighborhood.
This is done by testing all possible extension up to 2 genes for
all genes in the sub-graph. Again, this step is iteratively repeated until no
further genes are added to the significant differentially expressed
sub-graphs. This steps bridges small gaps of genes without a clear
differential signal in the DGEA~(Fig.~1c).\newline

\textbf{Merging Modulated sub-graphs}

After detection and extension of the modulated sub-graphs, they are
tested if combined sub-graph score is better than on their own. The merging of the two sub-graphs is done by depth first search, If the two sub-graphs merge with the connector of at most 2 genes and the
combined \textit{p}-value of the merged sub-graph including the
bridging genes in between is less than the individual
\textit{p}-values of the two sub-graphs, the two sub-graphs are merged
together to one big modulated sub-graph~(Fig.~1d). This step is repeated iteratively until no sub-graphs could be merged.\newline

@@

In a last post processing step \texttt{MSF} identifies the trigger
points of the modulated sub-graph. These trigger genes are the sources
of the sub-graphs with only outgoing edges. These genes can be
interpreted as the possible entry points of perturbation from where the
stimulus causes downstream effects. In the same spirit the most
downstream genes of the modulated sub-graph are identified and defined
as sinks. Sinks can be interpreted as the effectors where the
integrated information within the signal transduction network is set
to action.

@@

To demonstrate the usefulness \texttt{MSF} is applied to a RNAseq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus~\cite{Olejnik} (GSE84188). Ebola Virus (EBOV)
belongs to the Filoviridea family; filamentous, enveloped and single
stranded RNA viruses. EBOV causes hemorrhagic fever in humans,
inducing the host innate and adaptive immune response to be unable to
control virus infection~\cite{Prins}. Until now there are no approved
antiviral drugs for the treatment of Ebola virus infection~\cite{Konde,Rhein}. 
The initial targets of EBOV are the macrophages and
dendritic immune cells~\cite{Falasca,Rhein}. Ebola Virus inhibits the critical
innate immune response of the host, which includes the activation of
alpha/beta interferon (IFN-$\alpha /
\beta$)~\cite{Prins,Konde,Cardenas}. The EBOV viral protein VP35
targets the host type~I interferon IFN-$\alpha / \beta$ to block the
early antiviral immune response
~\cite{Prins,Konde,Falasca,Cardenas,Olejnik}. It has been proposed
that IFN-$\alpha / \beta$ could be used through clinical trials to
design antiviral drug against Ebola. The aim of testing the EBOV infection data with \texttt{MSF} was to  identify the modulated sub-graphs and check if \texttt{MSF} is able to identify the IFN-$\alpha / \beta$ gene as one of the sources.

The EBOV infection data has three time-points six hour (6hpi), one day (1dpi) and two day (2dpi) post infection. At 6hpi, five modulated sub-graphs were identified. Most of the genes part of the sub-moldules were chemokines (CXCL10, CCL8) and Interleukin genes (IL6, IL27, IL23) which serve as a subset of CD4+ T
cells. IFNB1  and IFNA1 were both identified as two of the sources. Most of the sources identified by \texttt{MSF} were type~I interferon induced genes. At 1dpi nine modulated sub-graphs were identified, IFNA1 was identified as one of the sources. For the last time-point again IFNB1 and IFNA1 were identified as the two sources out of the possible sources from seven sub-graphs identified.

As stated earlier IFN-$\alpha / \beta$ could be one of the target
genes of Ebola infection. We were able to successfully identify IFNA1
as a source in all three time-points and INFB1 in two of the time-points. Although IFNA1 and IFNB1 were two of the most
significant gene in the later time points, \texttt{MSF} was able to detect them as
a source in the early time-point when the genes were not significant
based on the individual DGEA alone. These sources will reduce the search space for potential target gene and can help the biologist as the starting point of clinical testing for drugs and vaccines against an infection.

\subsection*{Modulated sub-graphs at 6hpi}

The connected modulated sub-graphs from 6hpi are shown in~(Fig.~2). The sub-graph shows different colors for the predefined  KEGG pathways and how they are connected to each other. The RIG-like receptor signaling in connected to Toll-like receptor, which intern connects to MAPK signaling, JAK-STAT signalling, CYtokine-cytokine signaling, Chemokine receptor signaling and TNF signaling. These pathways could be seen further leading to WNT Signaling, P13K-Akt signaling.

From the other side the ECM receptor interaction is seen to lead to Focal adhesion which regulates the P13k-Akt signaling. Finally P13k-Akt signaling leads to FoxO signaling which eventually connects to cell cycle. TNF, P13k-Akt and FoxO signaling activates apoptosis and autophagy.

\subsection*{Comparison}

To better understand the functional role of the sub-graphs identified by \texttt{MSF}, we linked the  sub-graphs to the predefined KEGG pathways and Reactome
pathways through gene enrichment analysis. We used Cytoscape~\cite{Cyto} plugins to identify the enriched patwhays, CytoKegg~\cite{Cytokegg} for KEGG and Reactome FI~\cite{Reactome} for Reactome. We compared the pathways enriched in \texttt{MSF} identified sub-graphs to SPIA~\cite{Tarca} for KEGG and Reactome pathway enrichment analysis.

At 6hpi Cytokegg identified the sub-graph genes to be enriched in Toll-like receptor signaling, TNF signaling, IL-17 signaling and NF-kappa signaling pathways, as also stated by~\cite{Olejnik}. The earliest response EBOV induces on  cytokine is via TLR4-mediated signaling~\cite{Olejnik}, Gene enrichment analysis of sub-graph genes at 6hpi showed toll-like receptor signaling being most significantly dis-regulated pathway, this important signaling pathway was not identified by SPIA along with RIG-I-like receptor signaling pathway, TNF signaling pathway and IL-17 signaling (Table 1). 
On the later time-points \texttt{MSF} and SPIA showed consensus on the important pathways e.g Chemokine signaling, Cytokine-Cytokine receptor interaction, NF-kappa B signaling , Cytosolic DNA-sensing and RIG-I-like receptor signaling as being dis-regulated supplement (Table 3,4).

Likewise we compared gene enrichment results of sub-graph genes identified by \texttt{MSF} using Reactome FI to the Reactome predefined pathway enrichment analysis results. Although \texttt{MSF} and Reactome show good agreement on the dis-regulated pathways e.g Signaling by Interleukins, Interleukin-10 signaling, Cytokine Signaling in Immune system, Chemokine receptors bind chemokines, RIG-I/MDA5 mediated induction of IFN-alpha/beta pathways, NF-kB activation, Reactome as well fails to identify the Toll-like receptor signaling. This shows that \texttt{MSF} is able to identify the modulated genes in the sub-graphs at the earliest time-point of infection even though when the signal transduction was weak.

@@

To show the robustness and stability of our method, Poisson
distributed noise was added to the sample read counts. 
Then DGEA was carried out on the disturbed data with the
same parameters as for the native data, followed by analysis with
\texttt{MSF}. Using this criteria, the goal was to spot almost
identical modulated sub-graphs in the real and noisy data. This procedure was carried out 100 times and every time the genes from the sub-graphs identified from noisy data were compared to the sub-graphs identified from the native data. The robustness of \texttt{MSF} for the three time-points is shown in~(Fig.~2). For the last time-point (2dpi) 84\% genes of the sub-graphs from native data were retrieved, for 1dpi the robustness was about 79\%. The robustness was the lowest at 6hpi nearly 71\%, this could be due to that the infection merely started and the signal transdcution was weak as compared to the later time-points and over that Poisson distributed noise was added to the data.

@@

	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=10.0cm]{Git/MSF_Paper/Tgfb1} 
		\caption{}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=10.0cm]{Git/MSF_Paper/TGFBstep1}
		\caption{}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=10.0cm]{Git/MSF_Paper/TGFBstep2} 
		\caption{}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=10.0cm]{Git/MSF_Paper/GFBstep3}
		\caption{}
	\end{subfigure}
	\caption{Overview of \texttt{MSF} process showing the steps to identify the modulated sub-graphs. (a) showing a network of genes with green nodes as genes with \textit{p}-values $< $ 0.05, and blue nodes are genes with \textit{p}-values $>$ 0.05, (b) the genes circled red show the two initial modulated sub-graphs, sub-graph1 found one with M,K,C and the second modulated sub-graph2 S,V,T, (c) shows the modulated sub-graph1 being extended to M,K,C,A,D, (d) shows both the modulated sub=paths merged with the addition of 3 genes Z,W,P.}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[width=1.7\linewidth]{Git/MSF_Paper/Merged_Network}
	\caption{Modulated Sub-graph from 6hpi}
\end{figure*}

\begin{figure*}[p]
	\includegraphics[width=16cm]{Git/MSF_Paper/3Timepoints}
	\caption{Robustness analysis for three time-points of infection}
\end{figure*}
